Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-28 forecastflows.tex
Jim modifications

% Jim, if you want to add any of this into the abstract, please do!
%There's a fundamental problem in how a lot of empirically but observationally driven (i.e., non-experimental) research is done in ecology and related fields. The crux of the issue comes down to a couple of things. First, the research community has two very distinct paradigms for modeling the system or phenomenon or interest: either fitting simple trend-like models with only one or a few parameters, or developing forecast models that often include potentially complex, mechanistic, and/or black box submodels largely unconstrained by data. Second, regardless of the approach, these models tend to be developed not only in isolation of one another, but also without a coherent strategy for linking modeling/statistical decisions, data, and scientific understanding (both as input and output) throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development, incorporating sound statistical and scientific practices, and especially relying on data simulation to inform decisions at multiple steps in the process.

\begin{abstract}
Increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales. While new research often suggests that accounting for these complexities has a major impact on projected trends and forecasts, we argue that the typical approach to model fitting in ecology makes it impossible to evaluate and compare the models used to generate these insights and predictions. % jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated. Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater. We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights for both trend estimation and forecasting.
%Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
\end{abstract}

@@

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting. 
Instead forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, forecasting-focused process-based models often develop by adding new separate layers or components. These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and potential problems, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where and why different models diverge. Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection. %add cites for prereg?
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}, and most of macroecological insights emerge from exploring patterns in these datasets (exploratory data analysis).

@@

With a model in place the next step focuses on testing and understanding it via data simulation (step 3, fig. \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data. % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model. [FLAG: yes I can understand this point...]
% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. [FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage the focus is understanding the model, so researchers may have to spend a while to think through what parameters are reasonable or not. This is also a step for making sure the model is working as we expect, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data),

@@

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.

@@

In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% \todo[]{davies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set}

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another. % jdavies15april: is this more specific to process based  models? [FLAG: hmm yes again? Same point as above, maybe it's not clear we're talking mostly about PBM?]

@@

\columnbreak
\vspace*{1mm}
Blabla... \\
% We say here briefly how great the new approach is or something.
An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
Good model fitting is inherently iterative and takes time, at odds with the pressure to publish quickly.
Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).

@@

Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).

Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental % training 
separation in ecology has overly-siloed these groups and prevented more rapid progress.

Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting. 
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.

@@

In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made. Machine learning will likely surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning. % jdavies15april: nothing intrinsically bad about ML, but sometimes the goal is just to understand the system, not make predictions or forecasts.

% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development. As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.

>>>>

\begin{abstract}
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. While new research often suggests that models accounting for these complexities can yield more accurate trends and predictions, we argue that the usual workflows for model fitting in ecology make it impossible to evaluate and compare the models used to generate insights and predictions. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could be transformative for ecological modeling.

% jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated. Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater. We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights for both trend estimation and forecasting.
%Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
\end{abstract}

@@

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from massive geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components. These new parts are often disconnected from the original research aim, its data stream, and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).

Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge. Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses. This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection. %add cites for prereg?
But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}. Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).

@@

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, fig. \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data. % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model. [FLAG: yes I can understand this point...]
% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. [FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not. This is also a step for making sure the model is working as expected, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data). %jr24april not sure what this final sentence means?

@@

% More details, what's special! Feedbacks! %jrApr25 I think this paragraph reads really well!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand. Ecologists have started to use simulated data more to understand potential limitations of their models and data combined, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.

@@

In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results. % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave. pop decline since 1970. This would be massive!}
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% \todo[]{davies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set}

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another. % jdavies15april: is this more specific to process based  models? [FLAG: hmm yes again? Same point as above, maybe it's not clear we're talking mostly about PBM?]

@@

\columnbreak
\vspace*{1mm}
Blabla... \\
% We say here briefly how great the new approach is or something.
A universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
The first is pressure to publish quickly, which can be at odds with the reality that good model fitting is inherently iterative and takes time. The second is reluctance to embrace adoption of open science practices that ensure modeling efforts are fully transparent and reproducible. It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty. Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).

@@

Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).

Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Few of these groups have integrated data simulation into their statistical or scientific workflows. Simulation is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental % training
separation in ecology has overly-siloed these groups and prevented more rapid progress.

Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many others in the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.

@@

In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made. Machine learning will likely surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, engaging in statistical trend analysis, especially when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning. % jdavies15april: nothing intrinsically bad about ML, but sometimes the goal is just to understand the system, not make predictions or forecasts.

% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations. However, we believe our workflow provides an tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development. As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.
